---
title: "xCell2 Custom Prostate Reference"
output: html_document
date: "2026-04-20"
editor_options:
  chunk_output_type: console
---

## Goal

This notebook builds a custom xCell2 reference from `dge_combined_annotated.rds`.

The input object is a Seurat single-cell object. To make a stable custom reference for bulk RNA-seq deconvolution, this workflow creates pseudo-bulk reference profiles by averaging expression within each `refinedID` cell type and each source sample/dataset.

Reference single-cell: https://www.nature.com/articles/s41467-021-27322-4

## Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)

suppressPackageStartupMessages({
  library(data.table)
  library(Matrix)
  library(Seurat)
  library(xCell2)
  library(biomaRt)
  library(BiocParallel)
})

cfg <- list(
  seurat_rds = "/Users/YUC199/Library/CloudStorage/OneDrive-JohnsHopkins/Personal_data/Post-doc/Qian Lab JHU/Zhang Hui Lab/Hongyi/raw/dge_combined_annotated.rds",
  output_dir = "/Users/YUC199/Library/CloudStorage/OneDrive-JohnsHopkins/Personal_data/Post-doc/Qian Lab JHU/Zhang Hui Lab/Hongyi/raw/xCell2_custom_reference_tpm",
  assay = "RNA",
  expression_slot = "counts",
  celltype_col = "refinedID",
  dataset_col = "orig.ident",
  min_cells_per_pseudobulk = 20,
  min_pseudobulk_profiles_per_label = 2,
  gene_length_source = "Ensembl BioMart hsapiens_gene_ensembl transcript_length",
  train_ref_type = "rnaseq",
  use_ontology = TRUE,
  use_spillover = TRUE,
  spillover_alpha = 0.5,
  seed = 123,
  workers = 2
)

merge_reference_label <- function(x) {
  x <- gsub("_", "-", as.character(x))
  x[x %in% c("CD4-cluster1", "CD4-cluster2","CD8-cluster2","CD8-cluster3")] <- "T cell"
  x[x %in% c("NK-cell-cluster1", "NKcell-cluster2")] <- "T cell"
  x[x %in% c("Myeloid", "Proliferating Myeloid")] <- "T cell"
  x[x %in% c("CD14- Monocyte", "CD14+ Monocyte")] <- "Monocyte"
  x[x %in% c("M2-Macrophage", "M0-Macrophage")] <- "Macrophage"
  x
}

fetch_ensembl_gene_lengths <- function(genes, cache_file) {
  if (file.exists(cache_file)) {
    length_table <- fread(cache_file)
  } else {
    message("Fetching gene lengths from Ensembl BioMart. This is cached after the first run.")
    mart <- biomaRt::useEnsembl(
      biomart = "genes",
      dataset = "hsapiens_gene_ensembl",
      mirror = "useast"
    )
    gene_chunks <- split(unique(genes), ceiling(seq_along(unique(genes)) / 500))
    bm_list <- lapply(gene_chunks, function(gene_chunk) {
      biomaRt::getBM(
        attributes = c("hgnc_symbol", "ensembl_gene_id", "transcript_length"),
        filters = "hgnc_symbol",
        values = gene_chunk,
        mart = mart
      )
    })
    bm <- rbindlist(lapply(bm_list, as.data.table), fill = TRUE)
    length_table <- bm[
      !is.na(hgnc_symbol) & hgnc_symbol != "" &
        !is.na(transcript_length) & transcript_length > 0,
      .(gene_length_bp = max(transcript_length, na.rm = TRUE)),
      by = hgnc_symbol
    ]
    fwrite(length_table, cache_file, sep = "\t")
  }
  length_table
}

counts_to_tpm <- function(count_matrix, gene_length_bp) {
  gene_length_kb <- gene_length_bp / 1000
  rpk <- sweep(count_matrix, 1, gene_length_kb, "/")
  scaling_factor <- colSums(rpk, na.rm = TRUE) / 1e6
  sweep(rpk, 2, scaling_factor, "/")
}

dir.create(cfg$output_dir, recursive = TRUE, showWarnings = FALSE)
set.seed(cfg$seed)
```

## Load Seurat Reference

```{r load-reference}
ref_seurat <- readRDS(cfg$seurat_rds)
View(ref_seurat@meta.data)
stopifnot(inherits(ref_seurat, "Seurat"))
stopifnot(cfg$assay %in% names(ref_seurat@assays))
stopifnot(all(c(cfg$celltype_col, cfg$dataset_col) %in% colnames(ref_seurat@meta.data)))

Idents(ref_seurat) <- merge_reference_label(ref_seurat[[cfg$celltype_col, drop = TRUE]])
#Idents(ref_seurat) <- cfg$celltype_col

celltype_counts <- sort(table(merge_reference_label(ref_seurat[[cfg$celltype_col, drop = TRUE]])), decreasing = TRUE)
write.table(
  data.frame(label = names(celltype_counts), n_cells = as.integer(celltype_counts)),
  file = file.path(cfg$output_dir, "celltype_cell_counts.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

celltype_counts
```

## Build Pseudo-bulk Expression Matrix

For each `refinedID` x `orig.ident` group with at least `min_cells_per_pseudobulk` cells, raw counts are summed across cells to create pseudo-bulk profiles. Gene lengths are fetched from Ensembl BioMart and cached locally. Pseudo-bulk counts are then converted to TPM and transformed to `log1p(TPM)` for xCell2 training.

```{r build-pseudobulk}
expr_counts <- GetAssayData(ref_seurat, assay = cfg$assay, slot = cfg$expression_slot)
meta <- ref_seurat@meta.data
meta$cell_barcode <- rownames(meta)
meta$label_raw <- as.character(meta[[cfg$celltype_col]])
meta$dataset_raw <- as.character(meta[[cfg$dataset_col]])

meta$label <- merge_reference_label(meta$label_raw)
meta$dataset <- meta$dataset_raw
meta$pseudobulk_id <- paste(meta$label, meta$dataset, sep = "__")

group_counts <- as.data.table(meta)[
  ,
  .(n_cells = .N),
  by = .(label, dataset, pseudobulk_id)
]
group_counts <- group_counts[n_cells >= cfg$min_cells_per_pseudobulk]

label_profile_counts <- group_counts[, .(n_profiles = .N, total_cells = sum(n_cells)), by = label]
keep_labels <- label_profile_counts[n_profiles >= cfg$min_pseudobulk_profiles_per_label, label]
group_counts <- group_counts[label %in% keep_labels]

write.table(
  group_counts[order(label, dataset)],
  file = file.path(cfg$output_dir, "pseudobulk_groups_used.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

write.table(
  label_profile_counts[order(label)],
  file = file.path(cfg$output_dir, "label_pseudobulk_profile_counts.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

message("Pseudo-bulk groups retained: ", nrow(group_counts))
message("Labels retained: ", length(unique(group_counts$label)))

pseudobulk_list <- vector("list", nrow(group_counts))
for (i in seq_len(nrow(group_counts))) {
  group_id <- group_counts$pseudobulk_id[i]
  cells <- rownames(meta)[meta$pseudobulk_id == group_id]
  pseudobulk_list[[i]] <- Matrix::rowSums(expr_counts[, cells, drop = FALSE])
}

pseudobulk_counts <- do.call(cbind, pseudobulk_list)
colnames(pseudobulk_counts) <- make.unique(group_counts$pseudobulk_id)
rownames(pseudobulk_counts) <- rownames(expr_counts)
pseudobulk_counts <- as.matrix(pseudobulk_counts)

gene_length_cache <- file.path(cfg$output_dir, "ensembl_gene_lengths_by_hgnc_symbol.tsv")
gene_lengths <- fetch_ensembl_gene_lengths(rownames(pseudobulk_counts), gene_length_cache)
matched_length <- gene_lengths$gene_length_bp[match(rownames(pseudobulk_counts), gene_lengths$hgnc_symbol)]
has_length <- !is.na(matched_length) & matched_length > 0

write.table(
  data.frame(
    gene = rownames(pseudobulk_counts),
    has_ensembl_length = has_length,
    gene_length_bp = matched_length
  ),
  file = file.path(cfg$output_dir, "gene_length_match_summary.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

message("Genes with Ensembl length: ", sum(has_length), " / ", length(has_length))

pseudobulk_counts <- pseudobulk_counts[has_length, , drop = FALSE]
matched_length <- matched_length[has_length]

ref_matrix_tpm <- counts_to_tpm(pseudobulk_counts, matched_length)
ref_matrix <- log1p(ref_matrix_tpm)

keep_genes <- rowSums(is.finite(ref_matrix)) == ncol(ref_matrix) & matrixStats::rowVars(ref_matrix) > 0
ref_matrix <- ref_matrix[keep_genes, , drop = FALSE]
ref_matrix_tpm <- ref_matrix_tpm[rownames(ref_matrix), , drop = FALSE]

saveRDS(
  ref_matrix,
  file = file.path(cfg$output_dir, "prostate_xcell2_pseudobulk_log1p_tpm_matrix.rds")
)
saveRDS(
  ref_matrix_tpm,
  file = file.path(cfg$output_dir, "prostate_xcell2_pseudobulk_tpm_matrix.rds")
)
fwrite(
  data.table(gene = rownames(ref_matrix), as.data.frame(ref_matrix, check.names = FALSE)),
  file = file.path(cfg$output_dir, "prostate_xcell2_pseudobulk_log1p_tpm_matrix.tsv"),
  sep = "\t"
)
fwrite(
  data.table(gene = rownames(ref_matrix_tpm), as.data.frame(ref_matrix_tpm, check.names = FALSE)),
  file = file.path(cfg$output_dir, "prostate_xcell2_pseudobulk_tpm_matrix.tsv"),
  sep = "\t"
)

dim(ref_matrix)
```

## Build xCell2 Labels

The xCell2 labels table must include:

- `ont`: Cell ontology ID, or `NA` when not assigned.
- `label`: Cell type name.
- `sample`: Column name in the reference matrix.
- `dataset`: Source sample/dataset.

Ontology IDs are assigned only where the mapping is clear. Prostate cancer epithelial labels are intentionally left as `NA` because they are disease-state labels rather than standard normal cell ontology classes.

```{r labels}

group_counts$label%>%unique()
ontology_map <- c(
  "B-cell" = "CL:0000236",
  "Plasma" = "CL:0000786",
  "T-cell" = "CL:0000084",
  "CD8 T cell" = "CL:0000625",
  "CD4 T cell" = "CL:0000624",
  "Treg" = "CL:0000815",
  "NK-cell-cluster1" = "CL:0000623",
  "NKcell-cluster2" = "CL:0000623",
  "CD14+ Monocyte" = "CL:0001054",
  "CD14- Monocyte" = "CL:0000875",
  "M0-Macrophage" = "CL:0000235",
  "M1-Macrophage" = "CL:0000235",
  "M2-Macrophage" = "CL:0000235",
  "Myeloid" = "CL:0000763",
  "Proliferating Myeloid" = "CL:0000763",
  "Neutrophils" = "CL:0000775",
  "Eosinophils" = "CL:0000771",
  "Mast" = "CL:0000097",
  "Endothelial" = "CL:0000115",
  "Fibroblast" = "CL:0000057",
  "Smooth-muscle" = "CL:0000192",
  "Normal prostate epithelial" = "CL:0000066",
  "Club" = "CL:0000158"
)

labels <- data.frame(
  ont = unname(ontology_map[group_counts$label]),
  label = group_counts$label,
  sample = colnames(ref_matrix),
  dataset = group_counts$dataset,
  n_cells = group_counts$n_cells,
  stringsAsFactors = FALSE
)
labels$ont[is.na(labels$ont)] <- NA_character_

stopifnot(identical(labels$sample, colnames(ref_matrix)))
stopifnot(all(c("ont", "label", "sample", "dataset") %in% colnames(labels)))

write.table(
  labels,
  file = file.path(cfg$output_dir, "prostate_xcell2_reference_labels.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

table(labels$label)
```

## Optional Lineage File

This creates a lineage dependency file from available Cell Ontology IDs. You can inspect and manually edit the TSV before training if needed. Labels with `NA` ontology will be retained, but they will not contribute ontology relationships.

```{r lineage}
lineage_file <- file.path(cfg$output_dir, "prostate_xcell2_lineage_dependencies.tsv")
lineage_result <- tryCatch(
  xCell2GetLineage(labels = labels[, c("ont", "label", "sample", "dataset")], outFile = lineage_file),
  error = function(e) {
    message("Lineage generation failed; training will continue without lineage file: ", conditionMessage(e))
    NULL
  }
)

if (file.exists(lineage_file)) {
  lineage_file
} else {
  NA_character_
}
```

## Train xCell2 Custom Reference

```{r train-xcell2-reference}
parallel_param <- if (.Platform$OS.type == "unix" && cfg$workers > 1) {
  BiocParallel::MulticoreParam(workers = cfg$workers)
} else {
  BiocParallel::SerialParam()
}

set.seed(cfg$seed)

lineage_arg <- if (file.exists(lineage_file)) lineage_file else NULL

prostate_xcell2_ref <- xCell2Train(
  ref = ref_matrix,
  labels = labels[, c("ont", "label", "sample", "dataset")],
  refType = cfg$train_ref_type,
  useOntology = cfg$use_ontology,
  lineageFile = lineage_arg,
  BPPARAM = parallel_param,
  useSpillover = cfg$use_spillover,
  spilloverAlpha = cfg$spillover_alpha
)

saveRDS(
  prostate_xcell2_ref,
  file = file.path(cfg$output_dir, "prostate_custom_xCell2_reference.rds")
)

prostate_xcell2_ref
```

## Save Summary

```{r summary}
summary_lines <- c(
  "# xCell2 Custom Prostate Reference",
  "",
  paste("Input Seurat object:", cfg$seurat_rds),
  paste("Output directory:", cfg$output_dir),
  paste("Normalization:", "pseudo-bulk raw counts converted to TPM, then log1p(TPM) for xCell2Train"),
  paste("Gene length source:", cfg$gene_length_source),
  paste("Expression matrix dimensions:", paste(dim(ref_matrix), collapse = " x ")),
  paste("Pseudo-bulk profiles:", ncol(ref_matrix)),
  paste("Cell type labels:", length(unique(labels$label))),
  paste("Reference object:", file.path(cfg$output_dir, "prostate_custom_xCell2_reference.rds")),
  paste("Labels table:", file.path(cfg$output_dir, "prostate_xcell2_reference_labels.tsv")),
  paste("Expression matrix RDS:", file.path(cfg$output_dir, "prostate_xcell2_pseudobulk_log1p_tpm_matrix.rds")),
  paste("TPM matrix RDS:", file.path(cfg$output_dir, "prostate_xcell2_pseudobulk_tpm_matrix.rds")),
  paste("Lineage file:", ifelse(file.exists(lineage_file), lineage_file, "not generated")),
  "",
  "## Cell Type Counts",
  capture.output(print(table(labels$label)))
)

writeLines(summary_lines, file.path(cfg$output_dir, "README_xCell2_reference.md"))

sessionInfo()
```




## xcell
Raw enrichment scores: *RAW.txt, P-values for the significance of the scores to be different from zero: *pvals.txt.  In the manuscript we use a threshold of 0.2 to detemine a significant score. 

```{r}
#devtools::install_github('dviraran/xCell')
#



#fpkm <- fpkm %>%
#  dplyr::filter(!is.na(gene_name.T))%>%
#  dplyr::select(-gene_id.T, -gene_biotype.T,gene_biotype.N,gene_name.N) %>%   
#  group_by(gene_name.T) %>%
#  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
#  ungroup()

#fpkm <- fpkm %>%
#  column_to_rownames("gene_name.T")

#write.csv(fpkm,file = "../Output/RNAseq_fpkm.csv")
#devtools::install_github('AlmogAngel/xCell2')

library(xCell2)
library(readxl)
library(tidyverse)

tpm = read_excel("../Output/CPTAC_RNAseq_log2TPM.xlsx")%>%
  dplyr::filter(!is.na(gene_name.T))%>%
  dplyr::select(gene_name.T,contains("C3"),contains("CHTN"))%>% 
  group_by(gene_name.T) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
  ungroup()%>%
  column_to_rownames("gene_name.T")%>%
  as.matrix()

levels = c("Normal","No.Tumor","Low.Purity","multifocal","Sufficient.Purity")

meta = read.csv("../raw/raw/CPTAC/20_MetaData_03_12_2026.csv")

meta_RNA <- meta %>%
  filter(common_ID %in% colnames(tpm)) %>%
  column_to_rownames("common_ID") %>%
  mutate(condition = gsub("\\ ",".",FirstCategory)%>%
           factor(levels = group_levels))%>%
  dplyr::arrange(condition)

prostate_xcell2_ref = read_rds("/Users/YUC199/Library/CloudStorage/OneDrive-JohnsHopkins/Personal_data/Post-doc/Qian Lab JHU/Zhang Hui Lab/Hongyi/raw/xCell2_custom_reference_tpm/prostate_custom_xCell2_reference.rds")
xcell_res <- xCell2Analysis(
  mix = tpm,
  xcell2object = prostate_xcell2_ref,
  minSharedGenes = 0.8,spilloverAlpha = 0.8,rawScores =F
)


xcell_long = xcell_res%>%
  as.data.frame()%>%
  rownames_to_column("X")%>%
  reshape2::melt(id.vars = "X")%>%
  left_join(meta_RNA%>%
              rownames_to_column("variable"),by="variable")%>%
  mutate(condition = factor(condition,levels=group_levels))%>%
  dplyr::filter(!is.na(condition))%>%
  mutate(Celltype=ifelse(X=="Normal prostate epithelial","Normal Epithelial","Other cells"),
         Celltype=ifelse(X=="Tumor epithelial","Tumor Epithelial",Celltype),
         Celltype=ifelse(X%in%c("CD14+ Monocyte","M2-Macrophage","CD4 T cell","Mast","CD8 T cell","CD14- Monocyte","Myeloid","T-cell","M0-Macrophage" ),"Immune cells",Celltype),
         Celltype=ifelse(X%in%c("Endothelial","Fibroblast", "Club","Smooth-muscle"  ),"Stromal/Endo cells",Celltype))

df1 <- xcell_long %>% 
  group_by(X,condition) %>% 
  summarise(med = median(value))

ggplot(xcell_long%>%
         dplyr::filter(Celltype=="Immune cells")%>%
         dplyr::select(case_id,condition,value)%>%
         group_by(case_id,condition)%>%
         summarise(value=sum(value))%>%
         ungroup()%>%
         unique(),aes(x=condition,y=value+0.01)) +
  #geom_violin(scale = "width",trim = T)+
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                 geom = "crossbar", width = 0.6)+
  geom_dotplot(aes( fill = condition),color=NA,
                 binaxis = "y",         # which axis to bin along
                 binwidth = 0.01,        # Minimal difference considered diffeerent
                 stackdir = "center"    # Centered
                 )+
  #geom_point(position = position_jitterdodge(0.6),size=1)+
  theme_bw()+
  xlab("")+
  ylab("Log2 Enrichment Score")+
  theme(axis.text.x = element_blank())+
  scale_fill_manual(values = group_color)+
  theme(axis.text = element_text(size=11))+
  ggtitle("Immune cells (xCell2)")



ggplot(xcell_long%>%
         dplyr::filter(Celltype=="Tumor Epithelial")%>%
         dplyr::select(case_id,condition,value)%>%
         unique(),aes(x=condition,y=value)) +
  #geom_violin(scale = "width",trim = T)+
  geom_dotplot(aes( fill = condition),color=NA,
                 binaxis = "y",         # which axis to bin along
                 binwidth = 0.02,        # Minimal difference considered diffeerent
                 stackdir = "center"    # Centered
                 )+
  #geom_point(position = position_jitterdodge(0.6),size=1)+
  theme_bw()+
  xlab("")+
  ylab("Enrichment Score")+
  theme(axis.text.x = element_blank())+
  scale_fill_manual(values = group_color)+
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                 geom = "crossbar", width = 0.6)

write.csv(xcell_long,"../Output/xCell2_OutPut.csv")

Paired = xcell_long%>%
  reshape2::dcast(case_id+Celltype+X~Tissuetype,value.var = "value")%>%
  drop_na()%>%
  left_join(meta_RNA%>%
              dplyr::filter(!condition=="Normal")%>%
              dplyr::select(case_id,condition)%>%
              unique(),by="case_id")

library(ggrepel)
ggplot(Paired%>%
         dplyr::filter(Celltype=="Immune cells")%>%
         dplyr::select(case_id,condition,normal,tumor)%>%
         group_by(case_id,condition)%>%
         summarise(normal=sum(normal),
                   tumor=sum(tumor))%>%
         ungroup()%>%
         unique(),aes(x=normal,y=tumor,color=condition))+
  geom_point()+theme_bw()+
  geom_abline()+
  scale_color_manual(values = group_color)+
  scale_x_continuous(limits = c(0,0.5))+
  scale_y_continuous(limits = c(0,0.5))+
  geom_text_repel(data=Paired%>%
         dplyr::filter(Celltype=="Immune cells")%>%
         dplyr::select(case_id,condition,normal,tumor)%>%
         group_by(case_id,condition)%>%
         summarise(normal=sum(normal),
                   tumor=sum(tumor))%>%
         ungroup()%>%
         unique()%>%
           dplyr::filter(normal>tumor&condition=="No.Tumor"),aes(label=case_id),color="black")+
  theme(axis.text = element_text(size=11),
        axis.title = element_text(size=11))+
  xlab("NAT enrichment score")+
  ylab("Tumor enrichment score")+
  ggtitle("xCell2-Immune")

library(ggrepel)
ggplot(Paired%>%
         dplyr::filter(X=="Tumor epithelial"),aes(x=normal,y=tumor,color=condition))+
  geom_point()+theme_bw()+
  geom_abline()+
  scale_color_manual(values = group_color)+
  scale_x_continuous(limits = c(0,0.8))+
  scale_y_continuous(limits = c(0,0.8))+
  geom_text_repel(data=Paired%>%
         dplyr::filter(X=="Tumor epithelial")%>%
           dplyr::filter(tumor<normal),aes(label=case_id),color="black")+
  theme(axis.text = element_text(size=11),
        axis.title = element_text(size=11))+
  xlab("NAT enrichment score")+
  ylab("Tumor enrichment score")+
  ggtitle("xCell2")



mat <- xcell_long %>%
  dplyr::select(X, variable, value) %>%
  tidyr::pivot_wider(names_from = variable, values_from = value) %>%
  column_to_rownames("X") %>%
  as.matrix()
mat_z <- t(scale(t(mat)))   # row-wise z-score
mat_z[is.na(mat_z)] <- 0 
mat_z[mat_z > 2] <- 2
mat_z[mat_z < -2] <- -2
  
row_anno <- xcell_long %>%
  dplyr::select(X, Celltype) %>%
  distinct() %>%
  column_to_rownames("X")%>%
  mutate(Celltype=factor(Celltype,levels=c("Tumor Epithelial", "Normal Epithelial", 
                    "Immune cells", "Stromal/Endo cells")))%>%
  dplyr::arrange(Celltype)

col_anno <- meta_RNA %>%
  rownames_to_column("variable") %>%
  dplyr::select(variable, condition) %>%
  distinct() %>%
  column_to_rownames("variable")%>%
  dplyr::arrange(condition)

mat_z = mat_z[rownames(row_anno),rownames(col_anno)]

anno_color = list(
  "condition"= group_color
)
names(anno_color$condition) = group_level

library(pheatmap)

pheatmap(
  mat_z,
  cluster_cols = FALSE,   # no column clustering
  cluster_rows = FALSE,   # you already ordered rows
  annotation_col = col_anno,
  show_colnames = FALSE,
  annotation_colors = anno_color,
  show_rownames = TRUE,
  scale = "none",         # already scaled
  color = colorRampPalette(c("blue", "white", "red"))(100)
)

xcell_long_sum = xcell_long%>%
  group_by(condition,Celltype,variable)%>%
  summarise(value=sum(value))%>%
  ungroup()

ggplot(xcell_long_sum%>%
         dplyr::filter(Celltype=="Immune cells"),aes(x=condition,y=value,fill=condition))+
  geom_boxplot(outlier.shape = NA)+
  geom_point(color="black",position = position_jitter(0.2),alpha=0.4,size=1)+
  theme_bw()+
  facet_wrap(~Celltype,scales = "free_y")+
  xlab("")+
  ylab("Log2 (CPM +1)")+
  theme(axis.text.x = element_blank())+
  scale_fill_manual(values = group_color)+
  scale_y_continuous(limits = c(0,0.25))
  
LG_sufficient = mat%>%
  as.data.frame()%>%
  rownames_to_column("Celltype")%>%
  reshape2::melt(id.vars="Celltype")%>%
  group_by(variable)%>%
  mutate(sum = sum(value),
         perc = value/sum)%>%
  ungroup()%>%
  dplyr::filter(Celltype=="Tumor epithelial")%>%
  left_join(meta_RNA%>%
              rownames_to_column("variable"),by="variable")%>%
  mutate(Purity=as.numeric(Purity))%>%
  dplyr::filter(condition=="Sufficient.Purity")


library(ggpubr)
ggplot(LG_sufficient,aes(x=perc,y=Purity))+
  geom_point()+
  theme_bw()+
  xlab("xCell Purity")+
  ylab("Genomic Purity")+
  theme(axis.text = element_text(size=11))+
  scale_y_continuous(expand = c(0,0),limits = c(0,1),labels = scales::percent)+
  scale_x_continuous(expand = c(0,0),limits = c(0,1),labels = scales::percent)+
  geom_smooth(method = "lm", se = FALSE) +
  stat_cor(
    method = "pearson",
    label.x.npc = "left", 
    label.y.npc = "top"
  )


```

